---
title: "Assignment 3 - CRD"
author: "Ruben Vinueza"  
format:
  html:
    code-fold: False
    embed-resources: True
    toc: True
    number-sections: False
    theme: cerulean
---
# Instructions  
  - Move this script into the `code` subfolder of your `03_crd` project folder  
  - Move the file `wheat_nk_bamyan.csv` into the `data` folder of your `03_crd` project folder  
  - On this script, fill your name under the `title` section of the YAML  

# a) Setup  
Here is where we load the packages we will use.  
Make sure to load each package in the chunk below.  
```{r setup, warning=FALSE, message=FALSE}
library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function
library(broom) # for model residuals extraction
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display
```

Now, import the data set `wheat_nk_bamyan.csv` in the chunk below.  
```{r data import}
crd_a <- read_csv("../data/wheat_nk_bamyan.csv")

crd_a
```

The **objective of this study** was to assess the effect of different N and K rates on wheat yield.  

For that, a study was conducted using a completely randomized design as the experimental design with four replicates. The treatment design was a 3 N x 3 K rate factorial, for a total of 9 treatments.  


# b) EDA tables  
Use this space to conduct exploratory data analysis of the imported data, in the form of tables.  
```{r summary}
summary (crd_a)
```
```{r glimpse}
glimpse(crd_a)
```

# c) Wrangling
Here, make sure to wrangle your data set with all needed steps. Refer back to class code if you need a refresher on what was needed.  
```{r crd_aw1}
crd_aw1 <- crd_a %>%
  mutate(rep= factor(rep),
         nrate_kgha = factor(nrate_kgha),
         krate_kgha = factor(krate_kgha)
  ) %>%
  mutate (trtname = paste0(nrate_kgha,"+",krate_kgha))
         

crd_aw1
```
```{r crd_aw1}
summary(crd_aw1)
```
# d) EDA plots  
Now, explore the wrangled data using plots.  
Make sure to explore plots showing the main effects and the interaction.  
## Question 1) Below each plot of this section, write a sentence explaining your visual interpretation of the plot.  
```{r n boxplot}
ggplot(crd_aw1, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha
                    )) +
  geom_boxplot() +
  geom_jitter() +
 theme(legend.position = "none")
```
This graph shows that the more nitrogen rate in kg per ha used, the better yield of wheat we have. For example 200 kg of nitrogen gives as a median yield of ~5000 kg/ha of wheat. 
```{r k boxplot}
ggplot(crd_aw1, aes(x = krate_kgha, 
                    y = yield_kgha,
                    color = krate_kgha)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")
```
This graph shows that the more potassium rate in kg per ha used, the better yield of wheat we have as in the graph before of nitrogen. In this case using 60 kg/ha of potassium we have a yield of a median a little more than 5000 of yield of kg/ha of wheat

```{r nk boxplot}
ggplot(crd_aw1, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha
                    )) + 
  geom_boxplot() +
  geom_jitter() +
  facet_grid(.~krate_kgha)
```
In this plot just nitrogen alone as a fertilizer would not achieve the maximum yield as we can see in the first part of the graph. The middle part shows that 30 kg applied of potassium per ha with 200 kg of nitrogen show  the highest yields in kg per ha of wheat and the third part shows that 60 kg of potassium combined with 100 kg of nitrogen gives second biggest yield with median similar with using the maximum quantity of both fertilizers. 
# e) Statistical model  
In this section, run the appropriate model.  
Remember to set the contrast type to sum-to-zero (the default in R is set-to-zero, which is not what we want).  
Also, remember to specify all model terms needed for the treatment design.  

```{r model}
# Changing to sum-to-zero contrast
options(contrasts = c("contr.sum", "contr.poly"))

# Model fitting
crd_model <- lm(yield_kgha ~ nrate_kgha + krate_kgha + 
                nrate_kgha : krate_kgha,
              data = crd_aw1)
# Summary
summary(crd_model)
```

# f) ANOVA table  
Use this section to check the ANOVA table, ensuring you use type 3 sum-of-squares.
```{r ANOVA}
Anova(crd_model, type = 3)
```

## Question 2) Based on the ANOVA table above, for which effect(s) in the ANOVA should you extract means and perform pair-wise comparisons? Explain why.  
To the main factors or effects because they show statistical differences in the Anova table.

# g) Linear model assumptions  
Use this section to first compute your residuals, including studentdized residuals.  
## Question 3) What are the linear model assumptions that we need to check for? Create subsections (using ##) below for each of these assumptions, and write code to produce plots to check for these assumptions. For each assumption, write down your interpretation of the produced plots, and your thoughts if the assumptions hold or not.  
The linear model assumptions are based on the residuals which are assumed to be: mutually independent, normally distributed and do not have extreme observations or outliers. 
## Extracting residuals
```{r crd_residuals}
crd_residuals <- augment(crd_model) %>%
  mutate(.studresid = rstudent(crd_model))
 
crd_residuals
```
## Residual independence 
```{r}
ggplot(crd_residuals, aes( x = .fitted,
                       y = .studresid)) + 
  geom_point(shape = 21,
             fill = "green",
             size = 3,
             alpha = .7 
             ) + 
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth()
  
```
The residuals look independent they have not an obviuos pattern, progressive increment or quadratic shape. 
## Residual homoscedasticity 
```{r}
ggplot(crd_residuals, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "green", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```
It looks good. No shotgun effect. 
## Residual normality 
```{r}
ggplot(crd_residuals, aes(sample = .studresid)) +
  stat_qq() +
  stat_qq_line()
```
The residual points are close to the line with no tails so it falls a normal distribution in the Q-Q plot. 

```{r density plot}
ggplot(crd_residuals, aes(x = .studresid)) +
geom_density() +
scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3))
```
It still looks like a normal distribution. 

## Residual outliers  
```{r}
ggplot(crd_residuals, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "green", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```
It does not have any residual outliers. 

# h) Model means  
Use this section to extract model means according to your decision of the ANOVA table effects.  
```{r interaction means main factors}
crd_means_main <- emmeans (crd_model,
                          ~ nrate_kgha + krate_kgha
                         )
 
crd_means_main
```

# i) Pairwise comparisons  
Use this section to perform pair-wise comparison of the means extracted above.  

Make sure to reverse the order from high to low, use letters instead of numbers, and no adjustment.  
```{r interaction pwc main factors}
crd_cld_main <- cld(crd_means_main,
                   reverse = T,
                   Letters = letters,
                   adjust = "none"
                   )

crd_cld_main
```
```{r selected pwc}
crd_cld_select <- crd_cld_main %>%
  as.data.frame() %>%
  mutate(letter=trimws(.group)) %>%
  mutate(trtname=paste0(nrate_kgha, "+", krate_kgha))

crd_cld_select
```
# g) Final plot(s)  
Produce final plot(s) to show your results.  
Make sure to include both **raw data** and **statistical model summary (i.e., letter separation)**.  

Be mindful of your plot to ensure you are creating it in an intuitive way for your audience in relation to interpreting the letter separation structure you used.  

In class, we did not have time to make it publication-ready (e.g., a plot looking professional and ready to be included in a manuscript).

Here, make sure you deliver final plot(s) that is/are publication ready.  

Make sure to provide an interpretation for the plot(s) created, including which treatment levels were significantly higher/lower.

```{r crd final plot}
ggplot() +
  # Raw data and boxplots  
  geom_boxplot(data = crd_aw1, 
               aes(x = trtname, y = yield_kgha, fill = trtname), 
               width = 0.6, alpha = 0.7) + 
  geom_jitter(data = crd_aw1, aes(x = trtname, y = yield_kgha), 
              width = 0.2, alpha = 0.5, color = "black") +
  
  # Adding letters for pairwise comparisons  
  geom_label(data = crd_cld_select, aes(x = trtname, y = emmean + 200, label = letter), 
             size = 5, fontface = "bold", fill = "white", alpha = 0.8) +
  
  # Titles and axis labels  
  labs(title = "Effect of N and K Rates on Wheat Yield",
       x = "Treatment (N Rate + K Rate)", 
       y = "Yield (kg/ha)", 
       fill = "Treatment") +
  
  # Theme adjustments  
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x labels for readability
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "none")  # Hide legend since x-axis already shows treatments

```
The treatment that showed the lowest yield was 100 kg of nitrogen with no potassium while the treatments who gave the best yields in kg of wheat per ha were: 100 kg of nitrigen with 60 kg of potassium, 200 kg of nitrogen with 30 kg of potassium and 200 kg of nitrogen with 60 kg of potassium. 

# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `Assignment-03-crd-LASTNAME.html`.    

For ex., mine would be `Assignment-03-crd-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assigment #3 - CRD** by Feb 3rd 11:59 pm.  
  




